Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-26 paper.tex
Typo
Fixed a few notes

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}

@@

  with heterogeneous interfaces. \closenote{Marc-Etienne}{this -c2- can be also trigerred

@@

\emph{Scalability} corresponds to the ability of the architecture to
cope with high workloads. It is measured by counting the potential
scalability issues in the architectures, i.e. the lack of scalability
features. Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature could only be
implemented with a nonsensical amount of effort. Four different
features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): possibility to have
  more than 1 engine instance in the architecture. Workflow engines
  may require important amounts of resources when several workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to distribute the load among several
  engines. \texttt{S$_1$}=0 when adding a new engine instance is a
  fully automated process, i.e. the workflow engine only has to be
  started. In this case, elastic engines are possible, i.e. some kind
  of auto-scaling mechanism can be implemented to control the number
  of engine instances in the architecture. \texttt{S$_1$}=1 when
  adding a new engine instance requires some form of manual
  intervention (e.g., two-factor authentication), which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 when new engine
  instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): possibility to distribute
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 when distributed engines are possible in
  the architecture, and \texttt{S$_2$}=1 when they require specific
  developments in the workflow engine.\note{Rafael}{$S_2$ would be 2 if the paradigm does not allow to distribute the orchestration? e.g., workflow executions implemented within MPI jobs, in Pegasus PMC and dispel4Py.}
\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
  issue that depends more on the implementation of specific algorithms
  in the science gateway, workflow engine and infrastructure than on
  the architecture used to integrate the workflow engine in the
  science gateway. Some architectures, however, complexify the task
  scheduling problem by introducing additional software layers or
  create tasks with specific characteristics. \texttt{S$_3$}=0 when
  the architecture does not introduce any additional complexity to the
  scheduling problem, and \texttt{S$_3$}=1 otherwise.
\end{itemize}
The 3 metrics are summed to obtain a global measure of the scalability
issues that are present in the architecture.

@@

\paragraph{Scalability} Adding a new engine instance requires a new
instance of the science gateway, which is in general not possible
(\texttt{S$_1$}=2)\note{Tristan}{What-about load-balancing between web
  server instances?} \note{Marc-e}{Would that really be considered a real architectural feature or an IT hack?}.
  Distributed engines are not available by default
(\texttt{S$_2$}=1). The scheduling of tasks on the infrastructure is
as complex as in any other architecture since the workflow engine
might implement any kind of scheduling policy (\texttt{S$_3$}=0).

@@

\paragraph{Robustness} No component is specific to workflow execution
(\texttt{R$_1$}=0), and only interaction \texttt{e} is
(\texttt{R$_2$}=1)\note{sorina}{I understand why this is 1 (and why interactions
c and b are not counted here), but I wonder whether we can simply ignore these interactions
because they are not specific to workflow execution. As far as robustness is concerned,
the transfers have to be successful anyway...}.

@@

\paragraph{Integration effort} Setting up a nested workflow architecture with
service invocation requires a science gateway, 2 workflow services and
2 infrastructures (\texttt{I$_1$=5}). The architecture involves 11
interactions (\texttt{I$_2$=11}): \texttt{a$_1$} (appears twice: once for each
infrastructure), \texttt{a$_2$}, \texttt{b} (twice: once for each
infrastructure), \texttt{c$_1$} (twice: once for each infrastructure),
\texttt{c$_2$} (twice: once for each workflow engine and
infrastructure) and \texttt{d} (twice: once for each workflow engine).
\note{Rafael}{I like the idea of what has been discussed in the `integration effort', however I think the name may lead to misinterpretations. For example, according to Table 2, it requires more effort to integrate in a Tight architecture than in a service-based architecture. Considering the principles of software engineer, it is simpler to integrate/maintain a system using decoupled services. Maybe it should be renamed to something related to 'dependencies'.}

@@

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows,
which may be interesting in some cases.\note{Rafael}{I would suggest to provide some
more examples when this could be interesting.}
The availability of fine-grained debugging information may also be critical to efficient
user support.

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Overall, sub-tasking, and workflow import (with service invocation)
perform a bit better than tight integration, service invocation and
pool. Nested workflows stands a bit behind for the reasons mentioned
previously.\note{Rafael}{I would emphasize that this analysis is only for illustration purposes, since it does not account for different weights in metrics.}
\begin{table*}
\centering
\begin{tabular}{rcccccc}
                                    & \textbf{Tight}
                                    & \textbf{Service}
                                    & \textbf{Sub-task}
                                    & \textbf{Pool}
                                    & \textbf{Nested}
                                    & \textbf{Import} \\
  Integration effort (normalized total) &
                                    \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99E899}{0.22}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99D199}{0.44}
                                    & \cellcolor[HTML]{999999}{1.00}
                                      & \cellcolor[HTML]{99DD99}{0.33}\\
Robustness (normalized total) &
                                \cellcolor[HTML]{99F099}{0.14}
                                    & \cellcolor[HTML]{99D399}{0.43}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99B699}{0.71}
                                    & \cellcolor[HTML]{999999}{1.00}
                                    & \cellcolor[HTML]{99D399}{0.43}\\
  Extensibility (normalized total)  & \cellcolor[HTML]{99D199}{0.44}
                                     & \cellcolor[HTML]{99D199}{0.44}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99E899}{0.22}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99E899}{0.22}\\
Scalability (normalized total)  & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}\\
Specific features (normalized total) & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.00}\\
                                    & \cellcolor[HTML]{99E699}\textbf{2.1}
                                    & \cellcolor[HTML]{99E699}\textbf{2.1}
                                    & \cellcolor[HTML]{99FE99}\textbf{1.5}
                                    & \cellcolor[HTML]{99EE99}\textbf{1.9}
                                    & \cellcolor[HTML]{999999}\textbf{4.0}
                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}\\
\end{tabular}
\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
  are obtained by summing the normalized total scores ($m'$ values) of
  each criterion in Table~\ref{table:evaluation} and the colors are obtained by normalizing these scores
  as done in Table~\ref{table:evaluation}. }
\label{table:overall}
\end{table*}

>>>>

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\closednote}[4]{\pdfmargincomment[color=gray,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}

@@

  with heterogeneous interfaces. \closednote{Marc-Etienne}{this -c2- can be also trigerred

@@

\emph{Scalability} corresponds to the ability of the architecture to
cope with high workloads. It is measured by counting the potential
scalability issues in the architectures, i.e. the lack of scalability
features. Features are evaluated using a 3-level metric: 0 means that
the feature is very easy to enable, 1 means that it can be implemented
but with some difficulty, and 2 means that the feature could only be
implemented with a nonsensical amount of effort. Four different
features are identified:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Multiple engine instances (\texttt{S$_1$}): possibility to have
  more than 1 engine instance in the architecture. Workflow engines
  may require important amounts of resources when several workflows,
  or large workflows, are executed. At some point, it may be required
  for the science gateway to distribute the load among several
  engines. \texttt{S$_1$}=0 when adding a new engine instance is a
  fully automated process, i.e. the workflow engine only has to be
  started. In this case, elastic engines are possible, i.e. some kind
  of auto-scaling mechanism can be implemented to control the number
  of engine instances in the architecture. \texttt{S$_1$}=1 when
  adding a new engine instance requires some form of manual
  intervention (e.g., two-factor authentication), which prevents easy
  implementations of elastic engines. \texttt{S$_1$}=2 when new engine
  instances cannot be added.
\item Distributed engines (\texttt{S$_2$}): possibility to distribute
  the execution \emph{of a single workflow} among different engine
  instances. In our scope, this feature focuses on the capabilities of
  the architecture rather than these of the workflow
  engine. \texttt{S$_2$}=0 when distributed engines are possible in
  the architecture, \texttt{S$_2$}=1 when they require specific
  developments in the workflow engine..  \closednote{Rafael}{$S_2$
    would be 2 if the paradigm does not allow to distribute the
    orchestration? e.g., workflow executions implemented within MPI
    jobs, in Pegasus PMC and dispel4Py.}{Tristan}{The problem is that
    it is very difficult to define ``does not allow to''. For
    instance, why couldn't you hack Pegasus PMC so much that it
    becomes possible to distribute workflow executions in several MPI
    jobs coordinated by an MPI job? Moreover, S2=2 is not used so I'd
    prefer to leave it like this, unless you have a suggestion to
    reflect this in the text.}
\item Task scheduling (\texttt{S$_3$}): task scheduling is a difficult
  issue that depends more on the implementation of specific algorithms
  in the science gateway, workflow engine and infrastructure than on
  the architecture used to integrate the workflow engine in the
  science gateway. Some architectures, however, complexify the task
  scheduling problem by introducing additional software layers or
  create tasks with specific characteristics. \texttt{S$_3$}=0 when
  the architecture does not introduce any additional complexity to the
  scheduling problem, and \texttt{S$_3$}=1 otherwise.
\end{itemize}
The 3 metrics are summed to obtain a global measure of the scalability
issues that are present in the architecture.

@@

\paragraph{Scalability} Adding a new engine instance requires a new
instance of the science gateway, which is in general not possible
(\texttt{S$_1$}=2)\answerednote{Tristan}{What-about load-balancing between web
  server instances?}{Marc-e}{Would that really be considered a
  real architectural feature or an IT hack?}.  Distributed engines are
not available by default (\texttt{S$_2$}=1). The scheduling of tasks
on the infrastructure is as complex as in any other architecture since
the workflow engine might implement any kind of scheduling policy
(\texttt{S$_3$}=0).

@@

\paragraph{Robustness} No component is specific to workflow execution
(\texttt{R$_1$}=0), and only interaction \texttt{e} is
(\texttt{R$_2$}=1)\answerednote{sorina}{I understand why this is 1 (and why
  interactions c and b are not counted here), but I wonder whether we
  can simply ignore these interactions because they are not specific
  to workflow execution. As far as robustness is concerned, the
  transfers have to be successful anyway...}{Tristan}{We cannot just
  ignore these interactions in the architecture because they are
  required for the system to work. The transfers have to be successful
  but data movements in the workflow will use the same code as data
  movements done in the science gateway, which will make this
  architecture more robust than service for instance. In VIP, for
  instance, the code used in the GASW wrapper to transfer data (c2) is
  completely independent from the ``file transfer'' module in the
  platform (c1), which is detrimental to robustness}.

@@

\paragraph{Integration effort} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (appears twice: once for each infrastructure),
\texttt{a$_2$}, \texttt{b} (twice: once for each infrastructure),
\texttt{c$_1$} (twice: once for each infrastructure), \texttt{c$_2$}
(twice: once for each workflow engine and infrastructure) and
\texttt{d} (twice: once for each workflow engine).
\answerednote{Rafael}{I like the idea of what has been discussed in
  the `integration effort', however I think the name may lead to
  misinterpretations. For example, according to Table 2, it requires
  more effort to integrate in a Tight architecture than in a
  service-based architecture. Considering the principles of software
  engineer, it is simpler to integrate/maintain a system using
  decoupled services. Maybe it should be renamed to something related
  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
  simpler to integrate a service architecture than a tightly
  integrated one. But I take your point. 'Dependencies' is not very
  suitable either. What about 'development effort'?}

@@

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows,
which may be interesting in some cases. Also, sub-tasking and nested
workflows may complexify fine-grained debugging. \closednote{Rafael}{I would
  suggest to provide some more examples when this could be
  interesting.}{Tristan}{Done.}  The availability of fine-grained debugging
information may also be critical to efficient user support.

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Note that this analysis is only for illustration purposes since it
assumes that all criteria have equal weights while real systems would
favor some of them. Overall, sub-tasking, and workflow import (with
service invocation) perform a bit better than tight integration,
service invocation and pool. Nested workflows stands a bit behind for
the reasons mentioned previously.\closednote{Rafael}{I would emphasize
  that this analysis is only for illustration purposes, since it does
  not account for different weights in metrics.}{Tristan}{Done.}
  Integration effort (normalized total) &
                                    \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99E899}{0.22}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99D199}{0.44}
                                    & \cellcolor[HTML]{999999}{1.00}
                                      & \cellcolor[HTML]{99DD99}{0.33}\\
Robustness (normalized total) &
                                \cellcolor[HTML]{99F099}{0.14}
                                    & \cellcolor[HTML]{99D399}{0.43}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99B699}{0.71}
                                    & \cellcolor[HTML]{999999}{1.00}
                                    & \cellcolor[HTML]{99D399}{0.43}\\
  Extensibility (normalized total)  & \cellcolor[HTML]{99D199}{0.44}
                                     & \cellcolor[HTML]{99D199}{0.44}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99E899}{0.22}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99E899}{0.22}\\
Scalability (normalized total)  & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}\\
Specific features (normalized total) & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.00}\\
                                    & \cellcolor[HTML]{99E699}\textbf{2.1}
                                    & \cellcolor[HTML]{99E699}\textbf{2.1}
                                    & \cellcolor[HTML]{99FE99}\textbf{1.5}
                                    & \cellcolor[HTML]{99EE99}\textbf{1.9}
                                    & \cellcolor[HTML]{999999}\textbf{4.0}
                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}\\
\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
  are obtained by summing the normalized total scores ($m'$ values) of
  each criterion in Table~\ref{table:evaluation} and the colors are obtained by normalizing these scores
  as done in Table~\ref{table:evaluation}. }
\label{table:overall}
\end{table*}
